---
title: "Find best candidates for initial KD"
author: "Matt Lawlor"
date: "1/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# get interesting te-correlated genes

Here I take the results from male and female TLS LM results after correction.

```{r}
coefs <- bind_rows(female = read_tsv("subworkflows/dgrp_coex/results/linear_models/female_model_01/lm.tidy.corrected.tsv.gz"),
                   male = read_tsv("subworkflows/dgrp_coex/results/linear_models/male_model_01/lm.tidy.corrected.tsv.gz"),.id = "sex")

```

Here I get the universe of potentially interesting genes.

```{r}
# AnimalTFdb 3.0
tfs <- read_tsv("resources/Drosophila_melanogaster_TF.txt")
cofacs <- read_tsv("resources/Drosophila_melanogaster_TF_cofactors.txt")


tfoi <- bind_rows(tfs,cofacs) %>%
  dplyr::select(symbol=Symbol, feature.x = Ensembl,family = Family) %>%
  distinct()

symbol_lookup <- read_tsv("http://ftp.flybase.net/releases/current/precomputed_files/genes/fbgn_annotation_ID.tsv.gz", skip=3) %>%
  rename(symbol = "##gene_symbol", feature.x = "primary_FBgn#") %>%
  dplyr::select(symbol,feature.x)

coefs.oi <- filter(coefs, term == "x" & significant) %>%
  inner_join(symbol_lookup, by=c("feature.x"))
```

# filtering by dge in scrna

Now I want to filter the above list by DGE results. THis will take some processing. First filter for significant LFCs above a
relatively low threshold.

```{r}
dge <- list(head_female = "subworkflows/flycellatlas/results/downstream/integrated_by_tissue_and_sex_dge/head_female.dge.tsv.gz",
            head_male = "subworkflows/flycellatlas/results/downstream/integrated_by_tissue_and_sex_dge/head_male.dge.tsv.gz") %>%
  map_df(read_tsv,.id="dataset")

dge <- dge %>% filter(FDR < 0.1 & abs(summary.logFC) > 0.5)
```

I reason that the easiest, quickest filter will be to find genes that are up/downregulated in
clusters with TE up/downregulation.  I will do this by first identifying clusters that
seem highly amenable to TE expression or are highly repressive.

```{r}
top_groups <- dge %>% filter(!str_detect(feature,"FBgn")) %>%
  group_by(dataset,label, direction = ifelse(sign(summary.logFC)==1,"upregulated","downregulated")) %>%
  tally() %>%
  ungroup() %>%
  pivot_wider(names_from = direction,values_from = n) %>%
  replace_na(replace = list(upregulated=0,downregulated=0)) %>%
  mutate(ratio=log2((upregulated + 1)/(downregulated+1)))
```

We can visualize these groups here. I will only take the top few from each.

```{r}
plot_top_groups <- function(x) {
  mutate_at(x,"label",as.factor) %>%
  mutate(label = fct_reorder(label,ratio)) %>%
  mutate(ypos_dn = map_dbl(ratio,.f = function(x) {min(x + sign(x)*0.5,sign(x) * -0.5)})) %>%
  mutate(ypos_up = map_dbl(ratio,.f = function(x) {max(x + sign(x)*0.5,sign(x) * -0.5)})) %>%
    mutate(label=fct_drop(label)) %>%
  ggplot(aes(label,ratio)) +
  geom_col() +
  geom_text(aes(label, ypos_dn,label=downregulated), color="blue", size=rel(3)) +
    geom_text(aes(label, ypos_up,label=upregulated), color="red", size=rel(3)) +
  theme_classic() +
  theme(aspect.ratio = 0.5, axis.text.x = element_text(size=rel(1.2))) +
  facet_wrap(~dataset)  
}


top_groups %>% filter(dataset=="head_female") %>% plot_top_groups

top_groups %>% filter(dataset=="head_male") %>% plot_top_groups
```

Here I select the datasets/labels which I presume will have valid pairs.

```{r}
labels_to_select_from <- top_groups %>%
  group_by(dataset,enrichment = sign(ratio)) %>%
  slice_max(abs(ratio),n=5) %>%
  dplyr::select(dataset,label,enrichment) %>%
  filter(enrichment!=0) %>%
  ungroup()
```

Now is the slightly tricky data reshaping step - I want gene/te pairs
that are in a group with clear propensity for repression/activation.

I additionally filter to only include TEs that move in the overall direction of the group,
so that 1 or 2 TEs that are for example downrgulated in a permissive group don't contribute.

```{r}
# get all types of relationship that we want to search for.
intriguing_feature_sets0 <- dge %>%
  inner_join(labels_to_select_from, by = c("dataset","label")) %>%
  mutate(gene_direction = sign(summary.logFC)) %>%
  dplyr::select(dataset,label,feature,enrichment, gene_direction) %>%
  #filter(sign(summary.logFC)==enrichment) %>%
  mutate(type = ifelse(str_detect(feature,"FBgn"),"gene","TE")) %>%
  nest(-dataset,-label,-type,-enrichment,-gene_direction) %>%
  filter(!(type=="TE" & enrichment != gene_direction))

intriguing_feature_sets0
```

That last step means I need to add some TEs back.

```{r}
# now spread this, but replace the values for TEs.
intriguing_feature_sets1 <- intriguing_feature_sets0 %>%
  pivot_wider(names_from = type,values_from = data) %>%
  group_by(dataset,label) %>%
  mutate(TE = head(TE,1)) %>%
  ungroup()

intriguing_feature_sets1
```

Now I'll put the possible combinations into plain english.

```{r}
# get the relationships in plain english
intriguing_feature_sets2 <- intriguing_feature_sets1  %>%
    mutate(relationship_summary = case_when(enrichment == 1 & gene_direction == 1 ~ "pos",
                                            enrichment == 1 & gene_direction == -1 ~ "neg",
                                            enrichment == -1 & gene_direction == -1 ~ "pos",
                                            enrichment == -1 & gene_direction == 1 ~ "neg"))

intriguing_feature_sets2
```


```{r}
intriguing_feature_sets3 <- intriguing_feature_sets2 %>%
  mutate(features = map2(gene,TE, ~unlist(bind_rows(.x,.y)))) %>%
  dplyr::select(dataset,label,features, enrichment, gene_direction, relationship_summary)
```

```{r}
intriguing_feature_sets4 <- intriguing_feature_sets3 %>% unnest(features) %>% rename(feature = features)

filtering_feature_set <- inner_join(intriguing_feature_sets4 %>% filter(str_detect(feature,"FBgn")),
           intriguing_feature_sets4 %>% filter(!str_detect(feature,"FBgn")), by=c("dataset","label","enrichment","gene_direction","relationship_summary")) %>%
  mutate(sex = ifelse(str_detect(dataset,"female"),"female","male"))
```

```{r}

filtered_coefs <- coefs.oi %>%
  inner_join(filtering_feature_set,by=c(sex ="sex",feature.x="feature.x",feature.y="feature.y",relationship="relationship_summary")) %>%
  mutate(scrna_evidence = paste0(dataset,"_",label)) %>%
  group_by(feature.x, feature.y,estimate.qnorm,relationship,sex) %>%
  mutate(scrna_evidence = paste0(dataset,"_",scrna_evidence)) %>%
  summarise(n_supporting_clusters = n(), scrna_evidence = paste(scrna_evidence,collapse = ","),.groups = "drop") %>%
  distinct() %>%
  group_by(feature.x,relationship) %>%
  filter(n() > 3) %>%
  group_by(feature.x) %>%
  filter(length(unique(relationship))==1) %>%
  ungroup() %>%
  left_join(symbol_lookup)
```

```{r}
filtered_coefs %>%
  ggplot(aes(reorder(symbol,estimate.qnorm),estimate.qnorm)) +
  geom_jitter(width = 0.1) +
  geom_vline(xintercept = "pan") +
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust=1),aspect.ratio = 0.5) +
  facet_wrap(sex~relationship,scales = "free")
```

Ranking

```{r}
top.overall <- coefs.oi %>%
  group_by(sex, symbol) %>%
  summarise(mean.relationship = mean(estimate.qnorm),
            n.relationships=n()) %>%
  mutate(relationship = ifelse(sign(mean.relationship)==1,"pos","neg")) %>%
  group_by(sex,relationship) %>%
  mutate(rnk.effect = percent_rank(abs(mean.relationship)), rnk.n = percent_rank(n.relationships)) %>%
  ungroup()

tls_ranking <- top.overall %>%
  pivot_wider(names_from = sex, values_from = c(mean.relationship, n.relationships, relationship, rnk.n, rnk.effect)) %>%
  drop_na() %>%
  mutate(overall_score = rnk.n_male * rnk.n_female * rnk.effect_male * rnk.effect_female,
         male_score = rnk.n_male * rnk.effect_male,
         female_score = rnk.n_female * rnk.effect_female) %>%
  arrange(-overall_score) %>%
  relocate(overall_score,male_score,female_score,.after = symbol)
```


```{r}
# from TLS notebook on box
#tls_ranking <- read_tsv("~/Downloads/overall_ranking.tsv")

plottable_ranking <- filtered_coefs %>%
  group_by(symbol,relationship,sex) %>%
  summarise(n_tes = n(),scrna_evidence = paste(scrna_evidence,collapse = ","),.groups = "drop") %>%
  mutate(n_supporting_clusters = map_dbl(scrna_evidence,.f=function(x){length(unique(unlist(str_split(x,","))))})) %>%
  left_join(tls_ranking) %>%
  dplyr::select(symbol:female_score) %>%
  mutate(usable_score = ifelse(sex == "female",female_score,male_score)) %>%
  dplyr::select(-male_score,-female_score) %>%
  filter(n_tes > 3 & n_supporting_clusters > 1)

plottable_ranking %>%
  filter(n_tes > 3 & n_supporting_clusters > 1) %>%
  ggplot(aes(as_factor(n_tes),usable_score,label=symbol)) +
  geom_point(aes(size=n_supporting_clusters)) +
  facet_grid(. ~sex + relationship,scales = "free", space = "free") +
  ggrepel::geom_label_repel() +
  ylab("Rescaled Association Score") +
  theme_linedraw() +
  xlab("N supported TES")

plottable_ranking %>%
  filter(symbol%in% tfoi$symbol) %>%
  filter(n_tes > 3 & n_supporting_clusters > 1) %>%
  ggplot(aes(as_factor(n_tes),usable_score,label=symbol)) +
  geom_point(aes(size=n_supporting_clusters)) +
  facet_grid(. ~sex + relationship,scales = "free", space = "free") +
  ggrepel::geom_label_repel() +
  ylab("Rescaled Association Score") +
  theme_linedraw() +
  xlab("N supported TES")

plottable_ranking %>%
  filter(symbol%in% tfoi$symbol) %>%
  filter(n_tes > 3 & n_supporting_clusters > 1) %>%
  dplyr::select(symbol,n_tes,n_supporting_clusters,usable_score,sex) %>%
  distinct() %>%
  group_by(symbol) %>%
  mutate(n_tes = mean(n_tes), n_supporting_clusters = mean(n_supporting_clusters)) %>%
  ungroup() %>%
  pivot_wider(names_from = sex, values_from = c(usable_score), values_fill = 0) %>%
  ggplot(aes(female,male,size=n_tes,color=n_supporting_clusters, label=symbol)) +
  geom_point() +
  theme_linedraw() +
  ggrepel::geom_label_repel(size=rel(4))


plottable_ranking %>%
  dplyr::select(symbol,overall_score) %>%
  distinct() %>%
  arrange(-overall_score) %>%
  pull(symbol) %>%
  walk(message)
```


```{r}
sessionInfo()
```
